Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-29 numerical-ppm.tex

>>>>

\subsubsection{Riemann Solvers and Fallback}

While most numerical hydrodynamics schemes center around finding approximate
solutions to the hydrodynamics equations, the core of philosophy of a Godunov's
method \citep{Godunov1959} is to find the exact solution to an
approximation of the physical setup.  The approximate problem in this case is
the Riemann problem, wherein two volumes are separated by a thin membrane that
is instantaneously removed at time $t=0$.  The subsequent evolution has an exact 
analytic solution.  This solution is
described in detail in many texts on computational fluid dynamics
\citep[e.g.,][]{toro-1997}.   In brief, there are three waves that propagate
away from the initial discontinuity.  The central wave, characterized by a
density jump but not a pressure jump, is called the contact discontinuity.  The
waves traveling to the left and right of the contact discontinuity can be either
shocks, if characteristics converge on the wave front, or  rarefaction fans if
characteristics diverge.

While there exists an exact solution to this problem, finding it is expensive.  There are four
possible combination of left- and right- traveling shocks and rarefactions,
only one of which is fully consistent with the initial conditions.  Once the
correct physical state is determined, the pressure in the central region can
only be found by finding the root to an algebraic equation, which is necessarily
an iterative process.  Thus a series of \emph{approximate} Riemann solvers are
typically used.  There are four approximate Riemann solvers in Enzo: two-shock
\citep{toro-1997},
Harten-Lax-van Leer \citep[HLL,][]{toro-1997}, HLL with a contact discontinuity
 \citep[HLLC,][]{toro-1997}, and HLL with
multiple discontinuities \citep[HLLD,][]{Miyoshi05}.  Two-shock is used only
with the PPM method.  HLL and HLLC are used with PPM, MUSCL (both with and
without MHD) and MHD-CT.  HLLD is exclusively an MHD solver, and works with both
MUSCL and MHD-CT methods.

The only approximation that two-shock makes is that both left- and right- moving
waves are shocks.  This solution still requires an iterative method for finding
the pressure in between the two waves.  The HLL method alleviates this iteration
by assuming that there is no central contact discontinuity, and the signal speed
in the central region is approximated by an average over the left- and right-
moving waves.  This method is significantly faster than the two-shock method,
but also quite a bit more dissipative.  The HLLC is a three-wave method that improves upon the HLL
method by also including the third wave, the contact discontinuity.

For the MHD equations, there are seven waves, instead of three.  This makes the
exact solution to the Riemann quite a bit more expensive.  Both the HLL and HLLC
approximations can be formulated for the MHD equations, and are employed in both
Dedner and CT solvers in Enzo.  The HLLD
solver includes two of the additional waves, the rotational discontinuities,
making it a five-wave solver.

On rare occasions, high order
solutions can cause negative densities or energies.  Both PPM and MUSCL solvers employ a Riemann
solver fallback mechanism \citep{Lemaster09}.  If a negative density is found at
a particular interface, the more diffusive HLL Riemann solver is used instead.

### 2013-05-29 tests-fld-ifront.tex

This test problem examines Enzo's flux-limited diffusion radiation
transport and associated ionization chemistry solvers on a
cosmological I-front test.  This corresponds to test problem 4.6 from
\cite{ReynoldsHayesPaschosNorman2009}, run using the cosmological
deceleration parameter $q_0 = 0.05$ and initial redshift $z_i=10$.
Here, the physics of interest is the expansion of a \ion{H}{2} region
in a uniform gas around a single monochromatic ionizing source (with
frequency $h\nu = 13.6$ eV).  In this problem the I-front propagates
rapidly at first, approaching 90\% of the Str{\" o}mgren radius by a
scaled redshift of $-\log\left[(1+z)/(1+z_i)\right] \approx 0.15$,
before cosmological expansion overtakes ionization, pushing the Str{\"
o}mgren radius outward faster than the I-front can propagate.  Here,
the Str{\" o}mgren radius is analytically given by
\[
   r_S(t) = \left[\frac{3\dot{N}_{\gamma}}{4\pi \alpha_{\rm B}
   n_{\rm H}(t)^2}\right]^{1/3}, 
\]
where $\dot{N}_{\gamma} = 5 \times 10^{48}$ photon sec$^{-1}$ is the
stength of the ionizing source, $\alpha_{\rm B} = 2.52 \times 10^{-13}$
s$^{-1}$ is the \ion{H}{2} recombination rate, and $n_H(t)$ is the
proper number density of hydrogen.  We then define $\lambda =
\alpha_{\rm B}\, n_{H,i}\, /\, H_0\, / (1+z_i)$, where the subscript $i$
indicates the quantity at the initial redshift.  The analytical
solution for the I-front position as a function of time (or actually
of the time/redshift-dependent expansion coefficient $a(t)$) is then

\begin{eqnarray*}
   r_I(t) &=& r_{S,i} \left[\lambda e^{-\tau(a)} \int_1^{a(t)}
     e^{\tau(\tilde a)} \left(1 - 2q_0
     + \frac{2q_0(1+z_i)}{\tilde{a}}\right)^{-1/2}\mathrm
     d\tilde{a}\right]^{1/3}, \quad\text{where} \\ 
   \tau(a) &=& \lambda\frac{6q_0^2(1+z_i)^2}{F(a)-F(1)}, 
     \quad\text{and} \\
   F(a) &=& \left[2-4q_0 - \frac{2q_o(1+z_i)}{a}\right] 
      \left[1-2q_0 + \frac{2q_0(1+z_i)}{a}\right]^{1/2}.
\end{eqnarray*}

The simulation parameters are a box size $L_i\approx 27$ kpc, Hubble
constant $H_0 = 1$ (in units of 100 km/s/Mpc), and cosmological
constants $\Omega_m = 0.1$, $\Omega_A=0$ and $\Omega_b = 0.1$.  The
initial values for the simulation are a radiation energy density
$E_{r,i} = 10^{-35}$ erg cm$^{-3}$, temperature $T_i = 10^4$ K,
density $\rho_{b,i} = 2.35 \times 10^{-28}$ g cm$^{-3}$, and an
ionized hydrogen fraction $n_{\rm HII}/n_{\rm H} = 0$.

In Figure \ref{fig.fld}, we plot profiles of the \ion{H}{1} and
\ion{H}{2} fractions at the redshifts $z=\{6.24,\, 2.29,\, 1.02\}$, as
well as the ratio of the I-front and Str{\" o}mgren radii throughout
the simulation.  As seen in the left panel (species fractions as a
function of radius), the I-front is initially quite narrow, but slowly
becomes wider as the simulation proceeds due to the diffusion
approximation used by the radiative transfer scheme.  However, as seen
in the I-front propagation history plot (right panel), the computed
I-front location very accurately matches the exact value.

>>>>

This test problem examines \enzo's flux-limited diffusion radiation
transport and associated ionization chemistry solvers on a
cosmological I-front test.  This corresponds to test problem 4.6 from
\cite{ReynoldsHayesPaschosNorman2009}, run using the cosmological
deceleration parameter $q_0 = 0.05$ and initial redshift $z_i=10$.
Here the physics of interest is the expansion of an \ion{H}{2} region
in a uniform medium around a single monochromatic ionizing source
(with frequency $h\nu = 13.6$ eV).  The I-front propagates rapidly at
first, approaching 90\% of the Str{\" o}mgren radius by a scaled
redshift of $-\log\left[(1+z)/(1+z_i)\right] \approx 0.15$, before
cosmological expansion overtakes ionization, pushing the Str{\"
  o}mgren radius outward faster than the I-front can propagate.  The Str{\" o}mgren radius is analytically given by
\[
   r_S(t) = \left[\frac{3\dot{N}_{\gamma}}{4\pi \alpha_{\rm B}
   n_{\rm H}(t)^2}\right]^{1/3}, 
\]
where $\dot{N}_{\gamma} = 5 \times 10^{48}$ photon sec$^{-1}$ is the
strength of the ionizing source, $\alpha_{\rm B} = 2.52 \times 10^{-13}$
s$^{-1}$ is the \ion{H}{2} recombination rate, and $n_H(t)$ is the
proper number density of hydrogen.  We then define $\lambda =
\alpha_{\rm B}\, n_{H,i}\, /\, H_0\, / (1+z_i)$, where the subscript $i$
indicates quantities at the initial redshift.  The analytical
solution for the I-front position as a function of time is then

\begin{eqnarray*}
   r_I(t) &=& r_{S,i} \left[\lambda e^{-\tau(a)} \int_1^{a(t)}
     e^{\tau(\tilde a)} \left(1 - 2q_0
     + \frac{2q_0(1+z_i)}{\tilde{a}}\right)^{-1/2}\mathrm
     d\tilde{a}\right]^{1/3}, \quad\text{where} \\ 
   \tau(a) &=& \lambda\frac{6q_0^2(1+z_i)^2}{F(a)-F(1)}, \\
   F(a) &=& \left[2-4q_0 - \frac{2q_o(1+z_i)}{a}\right] 
      \left[1-2q_0 + \frac{2q_0(1+z_i)}{a}\right]^{1/2},
\end{eqnarray*}
and $a(t)$ is the cosmological expansion coefficient.

The simulation parameters are: box size $L_i\approx 27$ kpc, Hubble
constant $H_0 = 100$ km s$^{-1}$ Mpc$^{-1}$, and cosmological constants $\Omega_m = 0.1$, $\Omega_\Lambda=0$ and $\Omega_b = 0.1$.  The
initial values for the simulation are a radiation energy density
$E_{r,i} = 10^{-35}$ erg cm$^{-3}$, temperature $T_i = 10^4$ K,
density $\rho_{b,i} = 2.35 \times 10^{-28}$ g cm$^{-3}$, and an
ionized hydrogen fraction $n_{\rm HII}/n_{\rm H} = 0$.

In Figure \ref{fig.fld}, we plot profiles of the \ion{H}{1} and
\ion{H}{2} fractions at redshifts $z=\{6.24,\, 2.29,\, 1.02\}$, as
well as the ratio of the I-front and Str{\" o}mgren radii throughout
the simulation.  As seen in the left panel (species fractions as a
function of radius), the I-front is initially quite narrow, but slowly
becomes wider as the simulation proceeds due to the diffusion
approximation used by the radiative transfer scheme.  However, as seen
in the I-front propagation history plot (right panel), the computed
I-front location very accurately matches the exact value.

### 2013-05-29 tests-selfsiminfall.tex

The next test problem is based on the self-similar solutions to the cosmological spherical infall problem found by \citet{Bertschinger1985}.  It features a small density perturbation in a homogeneous $\Omega = 1$ universe and is a strong test of the cosmological evolution, gravity, and hydrodynamic portions of the code.  It is close analogue to cosmological cluster formation and has an analytic solution.

For initial conditions, we adopt a $32^3$ top grid with a single, initial sub grid (covering $1/8^3$ of the domain) with a refinement factor of 2.  An overdensity $\delta = 40$ is placed in a single cell near the center of the domain.  We begin at $z=199$ and evolve to $z=0$, which is a sufficient time for the evolution to largely forget its initial conditions and approach the self-similar result.  An over density refinement criteria ($\delta_{\rm crit} = 1.1$ on the top grid) is used to add additional grids, going up to 5 additional levels beyond the root grid.  We use the PPM solver without radiative cooling; only baryons are used for this problem.  An ideal gas law with $\gamma = 5/3$ is adopted.

In Figure~\ref{fig.sphericalinfall}, we show the results, scaled according to the dimensionless variables as defined in equations (2.9) and (3.2) of \citet{Bertschinger1985}.  This demonstrates that we can quickly and easily obtain a good solution with only a fairly modest initial grid.  The shock is sharply resolved and the asymptotic profiles as small $\lambda$ are recovered.  At very small values of $\lambda$, the initial conditions have not been fully forgotten and the self-similar solution is not recovered -- this can be seen most clearly in the dimensionless mass; however, this is simply because of the limited amount of time for which we evolved the solution.

\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{figures/SphericalInfall.eps}
\caption{The results of the self-similar spherical infall test.  The panels show, as a function of the dimensionless radius $\lambda$, the dimensionless density $D$ (top left), velocity $V$ (top right), pressure $P$ (bottom left), and enclose mass $M$ (bottom right).  In each case we show azimuthally averaged profiles from the simulation as circles and the analytic solution for for $\gamma = 5/3$ as solid lines.}
\label{fig.sphericalinfall}
\end{center}
\end{figure}

>>>>

This test problem is based on the self-similar solutions to the
cosmological spherical infall problem found
by \citet{Bertschinger1985}.  It features a small density perturbation
in a homogeneous $\Omega = 1$ universe and is a strong test of the
cosmological evolution, gravity, and hydrodynamic portions of the
code.  It is a close analogue to cosmological halo formation.

For the initial conditions, we adopt a $32^3$ top grid with a single,
initial sub grid (covering $1/8^3$ of the domain) with a refinement
factor of 2.  An overdensity $\delta = 40$ is placed in a single cell
near the center of the domain.  We begin at $z=199$ and evolve to
$z=0$, which is a sufficient time for the evolution to largely forget
its initial conditions and approach the self-similar result.  An
overdensity refinement criteria ($\delta_{\rm crit} = 1.1$ on the top
grid) is used to add additional grids, going up to 5 additional levels
beyond the root grid.  We use the PPM solver without radiative
cooling; only baryons are used for this problem.  An ideal gas law
with $\gamma = 5/3$ is adopted.

In Figure~\ref{fig.sphericalinfall}, we show the results, scaled
according to the dimensionless variables as defined in equations (2.9)
and (3.2) of \citet{Bertschinger1985}.  This demonstrates that we can
quickly and easily obtain a good solution with only a fairly modest
initial grid.  The shock is sharply resolved and the asymptotic
profiles at small $\lambda$ are recovered.  At very small values of
$\lambda$, the initial conditions have not been fully forgotten and
the self-similar solution is not recovered.  This can be seen most
clearly in the dimensionless mass.  However, this is simply because of
the limited amount of time for which we evolve the solution.

\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{figures/SphericalInfall.eps}
\caption{The results of the self-similar spherical infall test.  The
panels show, as a function of the dimensionless radius $\lambda$, the
dimensionless density $D$ (top left), velocity $V$ (top right),
pressure $P$ (bottom left), and enclosed mass $M$ (bottom right).  In
each case we show azimuthally averaged profiles from the simulation as
circles and the analytic solution for $\gamma = 5/3$ as solid lines.}
\label{fig.sphericalinfall}
\end{center}
\end{figure}

### 2013-05-29 tests-zeldovichpancake.tex

The Zel`dovich pancake \citep{1970A&A.....5...84Z} is particularly
relevant to cosmological simulations, because it includes many
features that are critical to structure formation: hydrodynamics,
expansion, and self gravity.  This problem represents the formation of an ideal,
isolated caustic, and is thus a useful proxy for much more
complicated structures in full 3-dimensional cosmological simulations
(such as the collapse of gas onto a cosmological halo or filament).

@@

with the internal energy of the gas set so that the entropy of the gas
maintains a constant value throughout.  The velocity perturbation is given by

@@

In the equations above, $\rho_0$ is the background density, z$_c$ is a
free parameter and is the redshift where the sheet forms a caustic
(i.e., where it `pancakes'), z is the redshift of initialization, x$_l$ is the
Lagrangian mass coordinate, $k = 2 \pi / \lambda$ (where $\lambda$ is
the perturbation wavelength), and H$_0$ is the $z = 0$ value of the
Hubble constant.  Note that this solution is expressed in terms of
Lagrangian positions, so one needs to convert this into the Eulerian
coordinates, x$_e$, that are more useful to \enzo:

@@

We note that the solution described above is exact up to the point of
caustic formation.  In Figure~\ref{fig.pancake}, we show the results
of a test of the adaptive mesh version of \enzo's Zel'dovich Pancake test.  A
one-dimensional box of length $64$~Mpc/h is initialized at z$ = 20$ in
an $\Omega_M = 1$ universe with h$ = 0.5$ and a background temperature
of 100~K, $\rho_0 = \rho_c$.  The simulation is initialized with 64
grid cells, refining by factors of four using criteria based on cell
mass and the presence of shocks, for a maximum of 2 levels (i.e., an
equivalent maximum resolution of 1024 grid cells).  The simulation was
evolved to z$ = 0$ using the PPM hydro method, with the final output
of the calculation being shown in Figure~\ref{fig.pancake}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/AMRZeldovichPancake.eps}
\caption{Zel`Dovich Pancake test shown at z$ = 0$, initialized in a
$\Omega_m = 1$ universe at z$ = 20$ on a one-dimensional grid having
64 cells, and further refined by factors of four based on cell mass
and the presence of shocks for up to two additional levels of mesh,
having a maximum effective resolution of 1024 grid cells. The top,
middle, and bottom rows show density, temperature, and velocity of the
gas, respectively; all are a function of position in units of the box
size.  The central region ($x \simeq 0.45-0.55$) has bene adaptively
refined, as can be seen by the locations of grid points.  Shocks and
the central density peak are clearly resolved, with well-delineated
jumps at the appropriate locations.}
\label{fig.pancake}
\end{center}
\end{figure}

>>>>

The Zel`dovich pancake \citep{1970A&A.....5...84Z} is particularly
relevant to cosmological simulations, because it includes many
features that are critical to structure formation: hydrodynamics,
expansion, and self gravity.  This problem represents the formation of an ideal,
isolated caustic, and is thus a useful proxy for much more
complicated structures in full 3-dimensional cosmological simulations,
such as the collapse of gas onto a cosmological halo or filament.

@@

with the internal energy of the gas set so that the entropy 
is constant throughout.  The velocity perturbation is given by

@@

In the equations above, $\rho_0$ is the background density, z$_c$ is a
free parameter and is the redshift where the sheet forms a caustic
(i.e., where it `pancakes'), z is the redshift of initialization, x$_l$ is the
Lagrangian mass coordinate, $k = 2 \pi / \lambda$ (where $\lambda$ is
the perturbation wavelength), and H$_0$ is the value of the Hubble
constant at $z = 0$.  Note that this solution is expressed in terms of 
Lagrangian positions, so one needs to convert this into the Eulerian
coordinates, x$_e$, that are more useful to \enzo:

@@

We note that the solution described above is exact up to the point of
caustic formation.  In Figure~\ref{fig.pancake}, we show the results
of a test of the adaptive mesh version of \enzo's Zel'dovich Pancake test.  A
one-dimensional box of length $64$~Mpc/h is initialized at $z = 20$ in
an $\Omega_M = 1$ universe with $h = 0.5$ and a background temperature
of 100~K, $\rho_0 = \rho_c$.  The simulation is initialized with 64
grid cells, refining by factors of four using criteria based on cell
mass and the presence of shocks, for a maximum of 2 levels (i.e., an
equivalent maximum resolution of 1024 grid cells).  The simulation is
evolved to z$ = 0$ using the PPM hydro method, with the final output
of the calculation shown in Figure~\ref{fig.pancake}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/AMRZeldovichPancake.eps}
\caption{Zel`Dovich Pancake test shown at z$ = 0$, initialized in a
$\Omega_m = 1$ universe at z$ = 20$ on a one-dimensional grid having
64 cells, and further refined by factors of four based on cell mass
and the presence of shocks for up to two additional levels of mesh,
having a maximum effective resolution of 1024 grid cells. The top,
middle, and bottom rows show density, temperature, and velocity of the
gas, respectively; all are a function of position in units of the box
size.  The central region ($x \simeq 0.45-0.55$) has been adaptively
refined, as can be seen by the locations of grid points.  Shocks and
the central density peak are clearly resolved, with well-delineated
jumps at the appropriate locations.}
\label{fig.pancake}
\end{center}
\end{figure}
